Programming Language Survey Paper on Haskell

### 2013-04-22 projects/term_paper/paper.tex
Finished the paper and added a few more bibliography entries.

\begin{minted}{haskell}
myExp =
    let e = 2.7182
    in  e ** x
\end{minted}

@@

\subsubsection{Iterative Constructs}
\bi
    \item List comprehensions
\begin{minted}{haskell}
filterModThree :: [Int] -> [Int]
filterModThree xs = [x | x <- xs, x `mod` 3 == 0]
\end{minted}
    \item Map/fold
\begin{minted}{haskell}
sumSquares :: (Num a) => [a] -> a
sumSquares xs = foldr1 (+) (map (**2) xs)
\end{minted}
    \item Recursion
\begin{minted}{haskell}
countdown :: Int -> String
countdown 0 = "Blast-off!"
countdown x = (show x) ++ "...\n" ++ countdown (x-1)
\end{minted}
\ei

\subsubsection{Functions}
\bi
    \item Definition/use
    \item Parameter passing techniques
\ei

\subsubsection{Scoping}
In Haskell, static scoping is employed.  As such, bindings are localized
to the function in which they were defined or passed.

\subsubsection{Modules}
\begin{minted}{haskell}
module myMod 
( 
addTwo,
addThree,
addFour
) where

@@

>>>>

\begin{minted}{haskell}
myExp :: (Num a) => a -> a
myExp =
    let e = 2.7182
    in  e ** x
\end{minted}

@@

\subsubsection{Iterative Constructs}

In Haskell, multiple constructs are supplied for the programmer to perform repetitive tasks including
list comprehensions, map- and fold-style functions, and function recursion.

List comprehensions in Haskell are a means to transform the elements in a list in some manner.
To facilitate the transformation, multiple parts are in the construct are used: the input elements, the
predicates, and the output function.  The iput elements are typically derived from another list.  Then,
the predicates are applied to each of these elements in other to restrict the input set.  Finally, the
restricted set then has the output function applied to every element in this set and the resuling list
is returned.  An exmaple below is given where \texttt{xs} is the input list, \texttt{x `mod` 3 == 0} is
a predicate used for filtering, and \texttt{x} is the simple output function.
\begin{minted}{haskell}
filterModThree :: [Int] -> [Int]
filterModThree xs = [x | x <- xs, x `mod` 3 == 0]
\end{minted}

Maps and folds are in many ways similar to list comprehensions.  In a map, a function is specified to apply to
every element of a list to produce a new list.  A fold in Haskell aggregates all of the elements in a list into
a single data type using some user-specified function.  An example using both map and reduce is given below.
In this example, every element in a list is first squared (mapped) and the resulting list is then summed up (folded).
\begin{minted}{haskell}
sumSquares :: (Num a) => [a] -> a
sumSquares xs = foldr1 (+) (map (**2) xs)
\end{minted}

All of the previous iterative constructs covered to this point share one thing in common:
they all could be implemented by recursion.  Recursion is the fundamental construct for iterating
in Haskell.  Recursion is implemented in functions in Haskell through pattern matching.  The boundary
case in a recursive definition terminates the iteration; the recursive process is typically defined using pattern matching,
where the boundary case is specificed with some pattern while the recurive case is set up as the default case.
The function below demonstrates this scenario by pattern matching on an integer value of the input parameter \texttt{x}.
\begin{minted}{haskell}
countdown :: Int -> String
countdown 0 = "Blast-off!"
countdown x = (show x) ++ "...\n" ++ countdown (x-1)
\end{minted}

It is worth noting that in Haskell, tail and guarded recursion are both utilized for improved performance 
by minimizing the function frame allocations on the stack.  However, because of the lazy-evaluation philosophy of Haskell,
recursions may be foregone until required, thereby negating the need for recursion optimization in some instances~\cite{haskell-tail-recur}.

\subsubsection{Functions}

Up to this point, several functions have been defined and used in the examples.  As Haskell is a functional programming
language, functions are the primary modular construct employed to accomplish tasks.  Worth noting here, the parameter passing techniques
supported in functions are by default explicit.  That is, bindings can be scoped to functions by explicitly invoking a function
with the arguments that want to be scoped to the function in general.  However, there are a few notable exceptions that the community
has developed to support implicit parameter passing to functions of some degree:
\bi
    \item Constant function definitions can be employed at the top level of a module~\cite{haskell-implicit-params}
    \item Monads can be utilized to effectively hide global definitions~\cite{haskell-implicit-params}
\ei

\subsubsection{Scoping}
In Haskell, static scoping is employed in resolving the origins of bindings.  As such, active bindings are first those localized
to the function followed by global scope.  As such, bindings within callee functions are not active in the caller function
unless explicitly passed as parameters.  This tenent of staticly binding within functions
works toward the end of keeping functions from producing any side-effects~\cite{haskell-scope}.

\subsubsection{Modules}

Modules in Haskell are the basic unit of organizing functions and other constructs.  The standard Haskell library
is composed of thousands of well-developed and maintained modules for a variety of purposes.  In the example below,
three functions are defined and made available to any other program that imports this module.

\begin{minted}{haskell}
module myMod 
( 
addTwo,
addThree,
addFour
) where

@@

In Haskell, there is a mature library of modules for built-in and extension-level
concurrency and parallelism~\cite{haskell-concurr}.  Platforms supported for concurrent and parallel execution include shared- and
distributed-memory systems, accelerator-style devices (e.g., general-purpose programming), and
symmetric and unsymmetric architectures~\cite{par-funk}.  Multi-threading and multiprocessing capabilities are available in
the standard Haskell library.  In the following example, multiple user-level threads are created for printing
to the main terminal device, one of which prints the word ``Hello'' while the other prints ``world.''

on graphics processing units)
\begin{minted}{haskell}
import Control.Concurrent
import System.IO

main :: IO ()
main = do { forkIO (hPutStr stdout "Hello")
            ; hPutStr stdout " world\n" }
\end{minted}
